---
title: "FA542 - Homework #3"
subtitle: "I pledge my honor that I have abided by the Stevens Honor System."
author: "Sid Bhatia"
date: "2023-10-17"
output: pdf_document
---

## Problem 1

Consider the weekly yields of Moody's Aaa and Baa seasoned bonds from January 5, 1962, to April 10, 2009 (*w-aaa.txt* and *w-Baa.txt* files).

The data are obtained from the Federal Reserve Bank of St Louis. Weekly yields are averages of daily yields.

```{r}
# Load libraries for skewness, kurtosis, and plotting.
library(moments)
library(fBasics)

# Establish the directory for data.
data_directory <- "C:/Users/sbhatia2/My Drive/University/Academics/Semester V/FA542 - Time Series with Application to Finance/FA542 - Homework/FA542 - Homework #3/FA542 - Homework #3 ~ Data/"

# Load in each dataset.
moody_Aaa <- read.table(paste(data_directory, 'w-aaa.txt', sep=""))
moody_Baa <- read.table(paste(data_directory, 'w-Baa.txt', sep=""))

head(moody_Aaa)
head(moody_Baa)

# Combine V1, V2, and V3 columns into a single date column.
moody_Aaa$Date <- as.Date(paste(moody_Aaa$V1, moody_Aaa$V2, moody_Aaa$V3, sep = "-"), format = "%Y-%m-%d")
moody_Baa$Date <- as.Date(paste(moody_Baa$V1, moody_Baa$V2, moody_Baa$V3, sep = "-"), format = "%Y-%m-%d")

# Keep only 'Date' and rename 'V4' to 'Aaa Bond Yield' and 'Baa Bond Yield' respectively.
moody_Aaa <- moody_Aaa[, c("Date", "V4")]
moody_Baa <- moody_Baa[, c("Date", "V4")]

colnames(moody_Aaa)[colnames(moody_Aaa) == "V4"] <- "Aaa Bond Yield"
colnames(moody_Baa)[colnames(moody_Baa) == "V4"] <- "Baa Bond Yield"

head(moody_Aaa)
head(moody_Baa)
```

#### a.

Obtain the summary statistics (sample mean, standard deviation, skewness, excess kurtosis, minimum, and maximum) of the two yield series. 

Are the bond yields skewed?

Do they have heavy tails? Answer the questions using 5% signifficance level.

```{r}
# Create function that calculates the sample mean, standard deviation, skewness, excess kurtosis, minimum, and maximum of simple returns.
compute_statistics <- function(returns)
{
  mean_val <- mean(returns)
  sd_val <- sd(returns)
  skewness_val <- skewness(returns)
  excess_kurtosis_val <- kurtosis(returns)
  min_val <- min(returns)
  max_val <- max(returns)
  
  # Create a list to hold each sample statistic.
  result <- list(
    mean = mean_val,
    standard_deviation = sd_val,
    skewness = skewness_val,
    excess_kurtosis = excess_kurtosis_val,
    minimum = min_val,
    maximum = max_val
  )

  return(result)
}

# Compute statistics for both bond yields.
compute_statistics(moody_Aaa$`Aaa Bond Yield`)
compute_statistics(moody_Baa$`Baa Bond Yield`)
```

We are testing the following:

$$H_0: S(r) = 0$$

$$H_a: S(r) \neq 0$$

Where $S(r)$ is the skewness of the bond yields.

The t-statistic for sample skewness is the following:

$$t = \frac{\hat{S}(r)}{\sqrt{6/n}}$$

```{r}
# Construct function to test whether the population skewness of yields is equal to 0 or not.
skewness_test <- function(yields)
{
  t_skewness <- skewness(yields) / sqrt(6 / length(yields))

  p_skewness <- 2 * (1 - pnorm(abs(t_skewness)))

  return(p_skewness)
}

# Conduct the skewness test for both yields.
skewness_test(moody_Aaa$`Aaa Bond Yield`)
skewness_test(moody_Baa$`Baa Bond Yield`)
```

Since both yields resulted in a p-value less than $0.05$, we reject the null hypothesis that the (population) skewness of the bond yields
are $0$, and we accept the alternative hypothesis that bond yields are skewed.

We are testing the following:

$$H_0: K = 3$$
$$H_a: K \neq 3$$

where $K$ denotes the kurtosis of the returns (excess kurtosis = 0).

The t-statistic for kurtosis of the yields is the following:

$$t = \frac{\hat{K}(r) - 3}{\sqrt{24/n}}$$

```{r}
# Construct function to test whether the population kurtosis of returns is equal to 3 or not (excess kurtosis = 0).
kurtosis_test <- function(yields)
{
  # Since the kurtosis function already computes excess log returns, we do not have to subtract by three.
  t_kurtosis <- (kurtosis(yields)) / sqrt(24 / length(yields))

  p_kurtosis <- 2 * (1 - pnorm(abs(t_kurtosis)))

  return(p_kurtosis)
}

# Conduct the kurtosis test for both yields.
kurtosis_test(moody_Aaa$`Aaa Bond Yield`)
kurtosis_test(moody_Baa$`Baa Bond Yield`)
```

Since both yields resulted in a p-value less than $0.05$, we reject the null hypothesis that the (population) kurtosis of the bond yields
are $3$, and we accept the alternative hypothesis that the kurtosis of the bond yields are not $3$, implying heavy tails.

#### b.

Build a time series model for the Aaa series.

```{r}
# Load in libraries for time series analysis.
library(tseries)

# Retrieve the Aaa bond yields.
aaa_bond_yield <- moody_Aaa$`Aaa Bond Yield`

# Convert Aaa bond yields to a time series object.
aaa_ts <- ts(aaa_bond_yield)

# Plot the time series.
plot(aaa_ts, main = "Aaa Bond Yield Time Series")

# Perform the ADF test to check for stationarity.
adf_test_result <- adf.test(aaa_ts)
adf_test_result
```

Since the p-value for the Augmented Dickey-Fuller (ADF) Test is greater than $0.05$, we fail to reject the null hypothesis that the 
time series data is non-stationary. As such, we will have to difference the data and test for stationarity.

```{r}
# Perform differencing.
differenced_aaa_ts <- diff(aaa_ts)

adf.test(differenced_aaa_ts)
```

Since the p-value of the ADF test is less than $0.05$, we can reject the null hypothesis that the difference time series data is 
non-stationary and accept the alternative hypothesis that the data is stationary at the $5%$ significance level.

```{r}
# Import the `forecast` library for ARIMA model.
library(forecast)

# Create a layout with one row and two columns.
par(mfrow = c(1, 2))

# Plot ACF and PACF for Aaa bond yields.
acf(aaa_ts)
pacf(aaa_ts)

# Perform grid search to find best ARIMA model based on AIC.
best_model <- NULL
best_aic <- Inf

for (p in 0:5) {
  for (d in 1:1) {
    for (q in 0:5) {
      model <- arima(aaa_ts, order = c(p, d, q))
      aic <- AIC(model)

      if (aic < best_aic) {
        best_model <- model
        best_aic <- aic
        best_order <- c(p, d, q)
      }
    }
  }
}

# Display best ARIMA model, AIC, and order from above.
best_model
best_aic
best_order

# Compare model with `auto.arima`.
auto.arima(aaa_ts)

auto.arima(aaa_ts)$aic < best_aic
```

As such, the best model for the Aaa bond yield data is $\text{ARIMA}(p = 4, d = 1, q = 5)$ since the model had a lower AIC than 
$\text{ARIMA}(p = 3, d = 1, q = 5)$.
