---
title: "FA542 - Homework #3"
subtitle: "I pledge my honor that I have abided by the Stevens Honor System."
author: "Sid Bhatia"
date: "2023-10-17"
output: pdf_document
---

## Problem 1

Consider the weekly yields of Moody's Aaa and Baa seasoned bonds from January 5, 1962, to April 10, 2009 (*w-aaa.txt* and *w-Baa.txt* files).

The data are obtained from the Federal Reserve Bank of St Louis. Weekly yields are averages of daily yields.

```{r}
# Load libraries for skewness, kurtosis, and plotting.
library(moments)
library(fBasics)

# Establish the directory for data.
data_directory <- "C:/Users/sbhatia2/My Drive/University/Academics/Semester V/FA542 - Time Series with Application to Finance/FA542 - Homework/FA542 - Homework #3/FA542 - Homework #3 ~ Data/"

# Load in each dataset.
moody_Aaa <- read.table(paste(data_directory, 'w-aaa.txt', sep=""))
moody_Baa <- read.table(paste(data_directory, 'w-Baa.txt', sep=""))

head(moody_Aaa)
head(moody_Baa)

# Combine V1, V2, and V3 columns into a single date column.
moody_Aaa$Date <- as.Date(paste(moody_Aaa$V1, moody_Aaa$V2, moody_Aaa$V3, sep = "-"), format = "%Y-%m-%d")
moody_Baa$Date <- as.Date(paste(moody_Baa$V1, moody_Baa$V2, moody_Baa$V3, sep = "-"), format = "%Y-%m-%d")

# Keep only 'Date' and rename 'V4' to 'Aaa Bond Yield' and 'Baa Bond Yield' respectively.
moody_Aaa <- moody_Aaa[, c("Date", "V4")]
moody_Baa <- moody_Baa[, c("Date", "V4")]

colnames(moody_Aaa)[colnames(moody_Aaa) == "V4"] <- "Aaa Bond Yield"
colnames(moody_Baa)[colnames(moody_Baa) == "V4"] <- "Baa Bond Yield"

head(moody_Aaa)
head(moody_Baa)
```

#### a.

Obtain the summary statistics (sample mean, standard deviation, skewness, excess kurtosis, minimum, and maximum) of the two yield series. 

Are the bond yields skewed?

Do they have heavy tails? Answer the questions using 5% signifficance level.

```{r}
# Create function that calculates the sample mean, standard deviation, skewness, excess kurtosis, minimum, and maximum of simple returns.
compute_statistics <- function(returns)
{
  mean_val <- mean(returns)
  sd_val <- sd(returns)
  skewness_val <- skewness(returns)
  excess_kurtosis_val <- kurtosis(returns)
  min_val <- min(returns)
  max_val <- max(returns)
  
  # Create a list to hold each sample statistic.
  result <- list(
    mean = mean_val,
    standard_deviation = sd_val,
    skewness = skewness_val,
    excess_kurtosis = excess_kurtosis_val,
    minimum = min_val,
    maximum = max_val
  )

  return(result)
}

# Compute statistics for both bond yield.
compute_statistics(moody_Aaa$`Aaa Bond Yield`)
```