---
title: "FA542 - Homework #1"
subtitle: "I pledge my honor that I have abided by the Stevens Honor System."
author: "Sid Bhatia"
date: "2023-09-16"
output: pdf_document
---

## Problem #1

### Import Libraries

```{r}
# Load libraries for skewness, kurtosis, and plotting.
library(moments)
library(fBasics)
```


### Data Retrieval

```{r}
# Establish the directory for data.
data_directory <- "C:/Users/sbhatia2/My Drive/University/Academics/Semester V/FA542 - Time Series with Application to Finance/FA542 - Homework/FA542 - Homework #1/HW1_data/"

# Load in each dataset.
data_problem_1 <- read.table(paste(data_directory, 'HW1_1.txt', sep=""), header = T)
data_problem_2 <- read.table(paste(data_directory, 'HW1_2.txt', sep=""), header = T)
data_problem_3 <- read.table(paste(data_directory, 'HW1_3.txt', sep=""), header = T)
```


#### 1a)

```{r}
df_1 <- as.data.frame(data_problem_1)

# Create function that calculates the sample mean, standard deviation, skewness, kurtosis, minimum, and maximum of simple returns.
compute_statistics <- function(returns)
{
  mean_val <- mean(returns)
  sd_val <- sd(returns)
  skewness_val <- skewness(returns)
  kurtosis_val <- kurtosis(returns)
  min_val <- min(returns)
  max_val <- max(returns)
  
  # Create a list to hold each sample statistic.
  result <- list(
    mean = mean_val,
    standard_deviation = sd_val,
    skewness = skewness_val,
    kurtosis = kurtosis_val,
    minimum = min_val,
    maximum = max_val
  )

  return(result)
}

# Compute statistics for each simple return series.
CAT_statistics <- compute_statistics(df_1$RET)
VW_statistics <- compute_statistics(df_1$vwretd)
EW_statistics <- compute_statistics(df_1$ewretd)
SP_statistics <- compute_statistics(df_1$sprtrn)

CAT_statistics
VW_statistics
EW_statistics
SP_statistics
```

#### 1b)

```{r}
# Display the empirical density distribution of simple returns for CAT.
hist(df_1$RET, nclass=30)

CAT_density_estimate <- density(df_1$RET)

# Obtain density estimate of simple returns for CAT.
CAT_density_estimate

# Perform a Jarque-Bera test for normality of simple returns.
normalTest(df_1$RET, method = 'jb')
```

At $\alpha = 0.05$ or at the 5% significance level, we reject the null hypothesis:
$$H_0: \text{The simple returns of CAT are normally distributed}$$ since the p-value
is less than 0.05.

#### 1c)

Log returns in relation to simple returns are defined as the following:

$$r_t = \ln(1 + R_t)$$

where $r_t$ are log returns and $R_t$ are simple returns.

```{r}
# Create function that transforms simple returns to log returns. log(...) is base e.
simple_to_log <- function(returns) {
  return(log(1 + returns))
}

# Convert simple returns to log using function.
CAT_log_returns <- simple_to_log(df_1$RET)
VW_log_returns <- simple_to_log(df_1$vwretd)
EW_log_returns <- simple_to_log(df_1$ewretd)
SP_log_returns <- simple_to_log(df_1$sprtrn)

# Use `compute_statistics()` function to calculate all relevant statistics.
CAT_log_statistics <- compute_statistics(CAT_log_returns)
VW_log_statistics <- compute_statistics(VW_log_returns)
EW_log_statistics <- compute_statistics(EW_log_returns)
SP_log_statistics <- compute_statistics(SP_log_returns)

CAT_log_statistics
VW_log_statistics
EW_log_statistics
SP_log_statistics
```

#### 1d)

To test the following the null hypothesis that the mean log returns are zero,
we need to conduct a t-test:

$$H_0: \text{Mean log returns are zero.}$$

$$H_0: \text{Mean log returns do not equal zero.}$$

```{r}
t.test(x = CAT_log_returns, alternative = c('two.sided'), mu = 0)
t.test(x = SP_log_returns, alternative = c('two.sided'), mu = 0)
```
As seen above, at the $\alpha = 0.05$ or the 5% significance level, we fail
to reject the null hypothesis that the mean log returns for both CAT and S&P
are zero since the p-values are well above 0.05.

#### 1e)

```{r}
# Display the empirical density distribution of log returns for CAT.
hist(CAT_log_returns, nclass=30)

CAT_log_density_estimate <- density(CAT_log_returns)

# Obtain density estimate of simple returns for CAT.
CAT_log_density_estimate

# Display the empirical density distribution of log returns for S&P
hist(SP_log_returns, nclass=30)

SP_log_density_estimate <- density(SP_log_returns)

# Obtain density estimate of simple returns for CAT.
SP_log_density_estimate
```

## Problem #2

#### 1a)

```{r}
df_2 <- as.data.frame(data_problem_2)

compute_statistics(df_2$RET)
compute_statistics(df_2$vwretd)
compute_statistics(df_2$ewretd)
compute_statistics(df_2$sprtrn)
```

#### 2b)

```{r}
# Display the empirical density distribution of simple returns for PG.
hist(df_2$RET, nclass=30)

PG_density_estimate <- density(df_2$RET)

# Obtain density estimate of simple returns for CAT.
PG_density_estimate

# Perform a Jarque-Bera test for normality of simple returns.
normalTest(df_2$RET, method = 'jb')
```

At $\alpha = 0.05$ or at the 5% significance level, we reject the null hypothesis:
$$H_0: \text{The simple returns of PG are normally distributed}$$ since the p-value
is less than 0.05.

#### 2c)

```{r}
compute_statistics(simple_to_log(df_2$RET))
compute_statistics(simple_to_log(df_2$vwretd))
compute_statistics(simple_to_log(df_2$ewretd))
compute_statistics(simple_to_log(df_2$sprtrn))
```
#### 2d)

To test the following the null hypothesis that the mean log returns are zero,
we need to conduct a t-test:

$$H_0: \text{Mean log returns are zero.}$$

$$H_0: \text{Mean log returns do not equal zero.}$$

```{r}
t.test(x = simple_to_log(df_2$RET), alternative = c('two.sided'), mu = 0)
t.test(x = simple_to_log(df_2$sprtrn), alternative = c('two.sided'), mu = 0)
```
As seen above, at the $\alpha = 0.05$ or the 5% significance level, we reject
the null hypothesis that the mean log returns for both CAT and S&P
are zero since the p-values below 0.05.

#### 2e)

```{r}
par(mfrow = c(1, 2))
hist(simple_to_log(df_2$RET), nclass=30)
hist(simple_to_log(df_2$sprtrn), nclass=30)

density(simple_to_log(df_2$RET))
density(simple_to_log(df_2$sprtrn))
```

## Problem 3

#### 3a)

```{r}
# Create function to calculate the confidence interval for a t-test.
t_confidence_interval <- function(alpha, n, mean, sd) 
{
  # Define degrees of freedom as n - 1.s
  df <- n - 1
  t_score <- qt(p = alpha/2, df = df)

  lower_bound <- mean - abs(t_score) * (sd / sqrt(n))
  upper_bound <- mean + abs(t_score) * (sd / sqrt(n))

  return(c(CAT_lower_bound, CAT_upper_bound))
}

# Define significance level alpha = 0.05 or 95% CI.
alpha <- 0.05

n <- length(CAT_log_returns)
CAT_mean <- CAT_log_statistics$mean
CAT_sd <- CAT_log_statistics$standard_deviation

t_confidence_interval(alpha, n, CAT_mean, CAT_sd)
```
#### 3b)

We are testing the following:

$$H_0: S(r) = 0$$

$$H_a: S(r) != 0$$

Where $S(r)$ is the skewness of the log returns.

The t-statistic for sample skewness is the following:

$$t = \frac{\hat{S}(r)}{\sqrt{6/n}}$$

```{r}
# Construct function to test whether the population skewness of returns is equal to 0 or not.
skewness_test <- function(returns)
{
  t_skewness <- skewness(returns) / sqrt(6 / length(returns))

  p_skewness <- 2 * (1 - pnorm(abs(t_skewness)))

  return(p_skewness)
}

skewness_test(CAT_log_returns)
```

Since the p-value is greater than 0.05, we fail to reject the null hypothesis
at the 5% significance level.

#### 3c)

We are testing the following:

$$H_0: K = 3$$
$$H_a: K != 3##

where $K$ denotes the kurtosis of the returns (excess kurtosis = 0).

The t-statistic for kurtosis of the log returns is the following:

$$t = \frac{\hat{K}(r) - 3}{\sqrt{24/n}}$$

```{r}
t_kurtosis <- (kurtosis(CAT_log_returns) - 3) / sqrt(24 / length(CAT_log_returns))

p_kurtosis <- 2 * (1 - pnorm(abs(t_kurtosis)))

p_kurtosis
```

Since the p-value is less than 0.05, we reject the null hypothesis at the 5% significance level.

## Problem 4

#### 4a)

We are testing the following:

$$H_0: S(r) = 0$$

$$H_a: S(r) != 0$$

Where $S(r)$ is the skewness of the daily log returns for S&P from 01/03/2007 to 12/31/2016.

```{r}
t_skewness <- skewness(SP_log_returns) / sqrt(6 / length(SP_log_returns))

p_skewness <- 2 * (1 - pnorm(abs(t_skewness)))

p_skewness
```

Since the p-value is less than 0.05, we reject the null hypothesis at the 5% significance level that the
log return is symmetric with respect to its mean.

#### 4b)

We are testing the following: 

We are testing the following:

$$H_0: K = 3$$
$$H_a: K != 3##

where $K$ denotes the kurtosis of the returns (excess kurtosis = 0).

```{r}
t_kurtosis <- (kurtosis(CAT_log_returns) - 3) / sqrt(24 / length(CAT_log_returns))

p_kurtosis <- 2 * (1 - pnorm(abs(t_kurtosis)))

p_kurtosis
```